Continous time Markov chain models; \\ extension, simulation and visualisation

### 2014-06-10 PhDThesis.tex
working on the last example

\subsection{Markov chain models of sequence substitution}
% TODO markov property is inherited from Poisson process
Drawing realizations with probabilities defined by $\mathbf{P}$  gives rise to a continuous-time stochastic process $\left\{ X(t):\; t\geq0\right\}$ satisfying the Markov property, such that for every $n\geq 0$, given the time points $0\leq t_{0}\leq t_{1}<\ldots<t_{n}\leq t_{n+1}$ and discrete states $i_{0},i_{1}, \ldots, i_{n},i_{n+1}$ it holds that:

@@

\ref{Jukes1969}

TODO

@@

\subsection{Bayes theorem}

@@

% TODO: better illustration, perhaps distances under some model of evolution?
\subsection{Example}

@@

\subsection{Marcov Chain Monte Carlo}

@@

%TODO: word for word, description
Metropolis-Hasting 
algorithm, as proposed by \citet{Metropolis1953}, is an MCMC algorithms for sampling from multi-dimensional distributions.
It proceeds by constructing a Markov chain with its states being the values of the parameter of interest $\theta$ and whose equilibrium distribution is the posterior $P\left(\theta|\mathbf{X}\right)$ as defined in Equation \ref{eq:bayesian1}.
After choosing an arbitrary starting state for the chain new candidate states $\theta^{*}$ are generated from the proposal density $q(\theta^{*} | \theta[t])$ also called the kernel.

@@

Figure \ref{fig:metropolis} shows $1000$ iterations of the chain generated by Algorithm \ref{alg:metropolisHastings} in the upper row.
The red line indicates the maximum likelihood estimate $\hat{\theta}$ of parameter $\theta$ which is calculated by setting $\frac{d\ell\left(\mathbf{X}|\theta\right)}{d\theta}=0$, where $l\left(\mathbf{X}|\theta\right)$ is the logarithm of likelihood defined by (\ref{eq:likelihood1}):

@@

In the second row of Figure \ref{fig:metropolis} we can see an estimate of the posterior density, along with the sample mean.
The acceptance rate is high, suggesting that almost as many steps are accepted as are rejected.

@@

>>>>

\subsection{Markov chain models of sequence substitution}

% TODO markov property is inherited from Poisson process
Drawing realizations with probabilities defined by $\mathbf{P}$  gives rise to a continuous-time stochastic process $\left\{ X(t):\; t\geq0\right\}$ satisfying the Markov property, such that for every $n\geq 0$, given the time points $0\leq t_{0}\leq t_{1}<\ldots<t_{n}\leq t_{n+1}$ and discrete states $i_{0},i_{1}, \ldots, i_{n},i_{n+1}$ it holds that:

@@

In this subsection we will discuss the simplest model of nucleotide substitution, i.e. the JC69 model \citep{Jukes1969}.
The JC69 model assumes that the instantaneous rates of change between every two nucleotides are the same.

@@

\subsection{Bayes theorem\label{sub:bayesTheorem}}

@@

\subsection{Example: Comparing two sequences with the beta-binomial model}

@@

\subsection{Marcov Chain Monte Carlo\label{sub:MCMC}}

@@

Metropolis-Hasting 
algorithm, as proposed by \citet{Metropolis1953}, is an MCMC algorithms for sampling from multi-dimensional distributions.
It proceeds by constructing a Markov chain with its states being the values of the parameter of interest $\theta$ and whose equilibrium distribution is the posterior $P\left(\theta|\mathbf{X}\right)$ as defined in Equation \ref{eq:bayesian1}.
After choosing an arbitrary starting state for the chain new candidate states $\theta^{*}$ are generated from the proposal density $q(\theta^{*} | \theta[t])$ also called the kernel.

@@

Figure \ref{fig:metropolis} shows $1000$ iterations of the chain generated by Algorithm \ref{alg:metropolisHastings} in the upper row and applied to the genetic distance problem.
The red line indicates the maximum likelihood estimate (MLE) $\hat{\theta}$ of parameter $\theta$ which is calculated by setting $\frac{d\ell\left(\mathbf{X}|\theta\right)}{d\theta}=0$, where $l\left(\mathbf{X}|\theta\right)$ is the logarithm of likelihood defined by (\ref{eq:likelihood1}):

@@

In the second row of Figure \ref{fig:metropolis} we can see an estimate of the posterior density, along with the sample mean.
We can see that the posterior sample mean is identical to the MLE up to the second decimal place, although we have not run the chain for long.
The acceptance rate is high, suggesting that almost as many steps are accepted as are rejected.

@@

Phylogeography is a research direction aimed at connecting evolutionary processes, which happen over time, with the processes of spatial distribution, in actual geographical coordinates into a joint spatio-temporal dynamic.
Historically however the phylogeographical inference in molecular epidemiology ignored the reciprocal character of these two processes. 
The evolutionary relationship represented by the tree was inferred omitting the geographical information, and then the ancestral locations generated by the process of the spatial diffusion were reconstructed, conditioning on the inferred tree topology.
\citet{Lemey2009} first considered the Bayesian reconstruction that allows for joint inference of both the tree and the parameters of the spatial and temporal process that shapes the spread and evolution of e.g. rapidly evolving viruses.
Their approach does not need to fix the tree topology, the branch lengths nor the Markov model parameters, and allows for integration over all possible values requiring only for a prior distribution specification for the parameters that we aim to infer.

%TODO: 
Central to the Bayesian phylogeographic framework is the hypothesis that sequence evolution is occurring simultaneously with the process of geographical dispersal. 
The observed data, residing at the $N$ tips of phylogeny $\mathbf{F}$, is recorded as both character sequence data $\mathbf{X}=(X_{1},...,X_{N})$ and discretized geographic locations (countries, districts, cities) $\mathbf{Y}=(Y_{1},...,Y_{N})$.
We assume independent stochastic processes are responsible for generating these data, with ancestral states of $\mathbf{X}$ generated as before by a CTMC characterized by rate matrix $\mathbf{Q}$ and the unobserved ancestral locations $(Y_{N+1},...,Y_{2N-2})$ generated by a separate CTMC characterized by rate matrix $\phi$.
The process is fully characterized by transition rates $\varphi_{ij}$ between states $i$ and $j$ and they occupy the elements of the transition rate matrix $\phi$.
Unlike before the rate matrix does not necessarily have to be symmetrical, i.e. there exist some $i,j$ such that $\varphi_{ij}\neq\varphi_{ji}$.
Additionaly, a priori one would expect many of the infinitesimal rates to be zero, as unlike the molecular data which typically is dense the location data is sparse, with only one observation per taxon and many connections not observed at all.

The Bayesian framework described in subsections \ref{sub:bayesTheorem} and \ref{sub:MCMC} offers the unique ability to integrate different sources of information about the viruses with their genetic data.
In the project we use this ability to infer the tree topology, the ancestral spatial locations and the character substitution process giving rise to the aligned molecular sequence data 
We assume independent stochastic processes are responsible for generating these data.
This allows us to write down the joint posterior distribution:

\begin{equation}
P\left(\mathbf{F},Q,\phi|\mathbf{X},\mathbf{Y}\right)\propto 
\underbrace{P(\mathbf{X}|\mathbf{F})}_{\begin{array}{c}
Likelihood
\end{array}}\cdot\underbrace{P(\mathbf{Y}|\mathbf{F})}_{\begin{array}{c}
Likelihood
\end{array}}\cdot 
\underbrace{P(\mathbf{F})}_{\begin{array}{c}
topology\\
prior
\end{array}}\cdot\underbrace{P(Q)}_{\begin{array}{c}
substitution\\
prior
\end{array}}\cdot\underbrace{P(\phi)}_{\begin{array}{c}
substitution\\
prior
\end{array}}
\label{eq:posteriorPhylogeography}
\end{equation}

The likelihood component for both molecular data $\mathbf{X}$ and location data $\mathbf{Y}$ is calculated using recursive tree pruning as descibed in subsection \ref{sub:likelihood}.
We approximate the joint posterior as written down in (\ref{eq:posteriorPhylogeography}) using MCMC methods we talked about in subsection \ref{sub:MCMC}.

%TODO: talk about the tree prior
